Assessing the Impact of MDE and Code Generation on the Sustainability of Scientific Computing Software: A Research Proposal

### 2019-09-17 ResearchProposal.tex
Updates to background and literature review section

Documentation can improve software sustainability.  For software
in general, documentation, written before and during development, provides many
benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
about requirements, more useful design reviews, easier integration of separately
written modules, more effective code inspection, more effective testing, and
more efficient corrections and improvements. For SCS in particular,
documentation provides benefits.  For instance, \citet{SmithEtAl2015-SS-TR}
shows that the quality of statistical software for psychology is generally
improved when developed using the structured CRAN (Comprehensive R Archive
Network) process and tools, versus an ad hoc process.
\citet{SmithAndKoothoor2016} highlights the value of proper documentation by
redeveloping nuclear safety analysis software.  Twenty seven (27) worrisome
documentation problems were found, including incompleteness, ambiguity,
inconsistency, verifiability, modifiability, traceability and a lack of
abstraction.  A redevelopment experiment with five existing projects
\citep{SmithJegatheesanAndKelly2016} enabled the code owners (as ascertained
through interviews) to clearly see the value of documentation.  However,
mirroring other studies \citep{CarverEtAl2007}, the code owners felt
documentation takes too much time.  Due to the time commitment and resource
limitation, documentation is rarely emphasized in SCS.

One promising approach to gain the benefits of documentation, but significantly
reduce the time commitment, is to use a Model Driven Engineering (MDE).  MDE is
an engineering approach that exploits structured representations (metamodels for
abstract syntax, and constraints to capture static semantics) to automate
repetitive, tedious and error prone tasks.  In this research proposal, we will
use MDE as a short-hand to include knowledge models, model transformation,
Domain Specific Languages (DSLs) and code generation.  Several current papers
point to DSLs and code/document generation as a transformative technology for
documentation, design and verification of SCS \citep{JohansonAndHasselbring2018,
  Smith2018}.  The benefit of an MDE approach is to dramatically reduce the cost
of developing reliable, reproducible and re-certifiable software.

One example of applying MDE to SCS is Drasil, which is introduced in
\citet{SzymczakEtAl2016} and available at
\href{https://github.com/JacquesCarette/Drasil}
{https://github.com/JacquesCarette/Drasil}.  Drasil is implemented as an open
source set of Domain Specific Languages (DSLs).  The Drasil approach involves:
i) creating infrastructure for a knowledge base of scientific and computing
models; and, ii) writing explicit ``recipes'' that weave together this knowledge
to generate theoretical models, design documents, code, test cases and build
scripts. The generator can render to multiple languages, such as html, LaTeX,
RTF for documentation and Python, Java, C, and Matlab, for code. One source of
knowledge, with rules for transformation, means completeness, consistency and
traceability can be achieved by construction. Moreover, these qualities can be
maintained as requirements are modified, design decisions are changed,
documentation standards are varied, and software is re­certified.  Although
creating the knowledge base is time consuming, the knowledge can be built up
incrementally.  New projects will reuse existing knowledge and expand as
necessary.

As described above, sustainability is a concern for SCS, and MDE holds promise
for addressing this concern.  However, a transformative change in the
development of SCS requires more than promise.  Empirical evidence is necessary
to demonstrate that an MDE process can improve the sustainability of SCS.  This
leads to the long-term objective for this research proposal: \emph{assess and
  measure the impact of MDE on the sustainability of SCS.}

The scope of SCS software for this proposal includes long-lived software, as
implied by the sustainability objective.  To gain the full benefits of
generating documentation, the scope also includes software that interests
multiple stakeholders, with different interests and backgrounds.  Finally, the
scope of this proposal emphasizes safety related software, such as software for
nuclear safety analysis, medical imaging and computational medicine.  Safety
related software is prioritized because documentation is often expected as part
of a certification exercise.  Although this study is not specifically restricted
to open-source software, for practical reasons, commercial software will not be
strongly emphasized.

Given the large number of potential MDE processes, techniques and technologies,
together with the wide variety in purpose, scope and context for SCS, the
long-term objective of measuring the impact of MDE is decomposed into four
short-term objectives.  For each objective, a reference is given to the
corresponding section in this document that provides the details.

\begin{enumerate}
\item Assess the current state of software development practice for SCS.  How is
  the software developed?  Are there existing development methods that lead to
  higher software sustainability? (Section~\ref{SecStateOfThePractice})
\item Assess the impact of an MDE process on end user developed SCS with respect
  to developer productivity and software sustainability.  
  (Section~\ref{SecImpactMDE})
\item Assess the impact on safety when the MDE generated documentation targets a
  safety assurance case.  An assurance case is an explicit structured argument
  pertaining to specific properties, such as trustworthiness.  The evidence for
  the argument will come from the traceability between artifacts generated by
  model transformations, expert reviews, test cases,
  etc. (Section~\ref{SecAssuranceCase})
\item Assess the impact on software quality when using MDE to facilitate
  Computational Variability Testing (CVT).  CVT uses code generation to build
  confidence in the generated code in an analogous way to the use of grid
  refinement studies.  Grid refinement looks at how the solution changes by
  varying the run-time parameter of grid density and comparing the results to
  theoretical expectations.  CVT, on the other hand, can generate code to
  ``refine'' build time parameters, such as order of interpolation, or degree of
  implicitness.  These parameters can be systematically varied and the results
  compared against the expected trend.  (Section~\ref{SecCVT})
\end{enumerate}

@@

To assess the impact of MDE on SCS quality, we need a clear definition of what
we mean by quality.  Section~\ref{SecSoftwareQuality} shows how the concept of
quality is decomposed into a set of separate qualities.  This set of qualities
can be applied to the software artifacts (documentation, test cases, etc) and to
the software development process itself.  Several of the qualities from the list
in Section~\ref{SecSoftwareQuality} cannot be measured directly, such as
maintainability.  Therefore, Section~\ref{SecDesirableQs} introduces measurable
documentation qualities that are believed to contribute to software quality.  In
some cases it may be necessary to use the qualities in
Section~\ref{SecDesirableQs} to indirectly measure qualities listed in
Section~\ref{SecSoftwareQuality}.

focuses on presents it, roadmap of this
section.  Some overview of the topics themselves, but an emphasis on how they
appear in the SCS context.

@@

\item [\textbf{Sustainability}] A combination of other qualities.

@@

To achieve the qualities listed in Section~\ref{SecSoftwareQuality}, the
documentation should achieve the qualities listed in this section.  All but the
final quality listed (abstraction), are adapted from the IEEE recommended
practise for producing good software requirements~\cite{IEEE1998}.

@@

Although reproducibility is the cornerstone of the scientific method, until
recently it has not been treated seriously in software
\citep{BenureauAndRougier2017}.  Fortunately, in recent years multiple
conferences, workshops and individuals are calling for dramatic change
\citep{BaileyEtAl2016}.  The need for action is highlighted by a study of 402
computer systems papers - only 48.3\% of the code was both available and
compilable \citep{CollbergEtAl2015}.  (Drasil addresses this problem because as
programming languages evolve the code renderers in Drasil can be updated.)
Reproducibility problems are even more extreme when the goal is replicability.
A third party should be able to repeat a study using only the description of the
methodology from a published article %, with no access to the original code or
%computing environment 
\citep{BenureauAndRougier2017}.  However, replicability is rarely achieved, as
shown for microarray gene expression \citep{IoannidisEtAl2009} and for economics
modelling \citep{IonescuAndJansson2013}.  Drasil addresses completeness and
ambiguity problems, since it emphasizes capturing and documenting all of the
required knowledge, including derivation of equations and
rationales. \citet{CrickAndHall2014} point out potential roadblocks for
reproducibility, including page length constraints and differing detail needs
depending on the audience.  Again Drasil addresses these concerns because the
recipes used to generate the documentation can be tailored to the level of
detail required.

@@

Thankfully SCS leaders recognize that an interdisciplinary approach provides the
path forward.  They believe that the solution to SCS quality problems is
applying, adapting and developing SE methods, tools and techniques.  However,
typical software processes are a barrier to progress.  ``To break the gridlock,
we must establish a degree of cooperation and collaboration with the [SE]
community that does not yet exist'' \citep{FaulkEtAl2009}.  ``There is a need to
improve the transfer of existing practices and tools from ... [SE] to scientific
programming. In addition, ... there is a need for research to specifically
develop methods and tools that are tailored to the domain'' \citep{Storer2017}.
\emph{This tailored research will require individuals, like myself, that have a
  multidisciplinary background.}

@@

To achieve the qualities listed in Table~\ref{Tbl_Qualities}, the documentation
in Table~\ref{TblDocuments} should have the following qualities:
\emph{complete}, \emph{correct}, \emph{consistent}, \emph{modifiable},
\emph{traceable}, \emph{unambiguous}, and \emph{verifiable}.  All of these
qualities are listed in the IEEE recommended practice for software
requirements~\cite{IEEE1998}.  The IEEE guidelines are for requirements, but
most qualities are relevant for all documentation artifacts.  Another relevant
quality, which is not on the IEEE list, is \emph{abstract}.  Requirements should
state what is to be achieved, but be silent on how it is to be achieved.
Abstraction is an important software development principle for dealing with
complexity~\cite[p.~40]{GhezziEtAl2003}.  Smith and Koothoor present further
details on the qualities of documentation for scientific
software~\cite{SmithAndKoothoor2016}.

@@

DSLs and code/document generation provide a transformative technology for
documentation, design and verification \citep{JohansonAndHasselbring2018,
  Smith2018}.  DSLs allow scientists their preferred approach of focusing on
science not software \citep{Kelly2007}.  A generative approach removes the
maintenance nightmare of documentation duplicates and near duplicates
\citep{LucivEtAl2018}, since knowledge is only captured once and automatically
transformed as needed.  Code generation has previously been applied to improve
SCS \citep{WhaleyEtAl2001, Veldhuizen1998, Pueschel2001}.  For instance, ATLAS
(Automatically Tuned Linear Algebra Software) \citep{WhaleyEtAl2001} and Blitz++
\citep{Veldhuizen1998} produces efficient and portable linear algebra software.
Spiral \citep{Pueschel2001} uses software/hardware generation for digital signal
processing.  \citet{Carette2008} shows how to generate a family of efficient,
type-safe Gaussian elimination algorithms.  FEniCS (Finite Element and
Computational Software) \citep{LoggEtAl2012} uses code generation when solving
differential equations.  Unlike previous work on SCS code generation, Drasil
will focus on generating all software artifacts (requirements, design etc.), not
just code.

\citet{SzymczakEtAl2016} removes excuses for avoiding documentation by providing
transformative SCS development technology.  Drasil provides an infrastructure
for knowledge capture and document/code generation.  Jacques Carette and I,
along with our students, have developed a prototype (available on GitHub),
implemented via Domain Specific Languages (DSLs) embedded in Haskell.  Drasil
already generates requirements documentation and code for several case studies,
including simulating the temperature of a solar water heating system, glass
breakage and two-dimensional game physics.  The full documentation
(requirements, design etc), code and test cases have been created manually for
each case study.  These manual case studies (many from \citet{SmithJegatheesanAndKelly2016})
provide the `gold standard' against which Drasil is tested.  All of these
artifacts are available publicly.  The generation techniques for Drasil began to
take shape during my work on generating geometric data structures and functions
\citep{CaretteEtAl2011}.

@@

empirical software engineering paper, case study paper

@@

@@

Scientists generally use an approximation of agile methods for their software
development.  In addition, documentation is generally not emphasized.  How
should Drasil be fit into the scientific software development process?  The
Drasil process should encourage an early intensity of thought and effort, but at
the same time, it shouldn't require waterfall development.  Drasil supports
frequent chance and incremental development, but where should developers start
and how should the work proceed?  How does one create an almost empty project in
Drasil and then borrow from existing knowledge and create new knowledge?  This
project could start with something like the task described in~\ref{SecPDF}.

@@

Experimentally Determine Appropriate Documentation Template: The best approach
for documenting requirements and design is an open question.  Drasil could be
used to attempt to address this question.  Different recipes for documentation
could be developed and compared.  Some comparison could be automated between the
documents, such as empirical measures of their quality
(Section~\ref{SecEmpiricalMeasureDoc}).  Other measures would require some kind
of usability experiments.

>>>>

Documentation can improve software sustainability.  For software in general,
documentation, written before and during development, provides many
benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
about requirements, more useful design reviews, easier integration of separately
written modules, more effective code inspection, more effective testing, and
more efficient corrections and improvements. For SCS in particular,
documentation provides significant benefits.  For instance,
\citet{SmithEtAl2015-SS-TR} shows that the quality of statistical software for
psychology is generally improved when developed using the structured CRAN
(Comprehensive R Archive Network) process and tools, versus an ad hoc approach.
\citet{SmithAndKoothoor2016} highlights the value of proper documentation by
redeveloping nuclear safety analysis software.  Twenty seven (27) worrisome
documentation problems were found, including problems with incompleteness,
ambiguity, inconsistency, verifiability, modifiability, traceability and a lack
of abstraction.  A redevelopment experiment with five existing projects
\citep{SmithJegatheesanAndKelly2016} enabled the code owners (as ascertained
through interviews) to clearly see the value of documentation.  However,
mirroring other studies \citep{CarverEtAl2007}, the code owners felt
documentation takes too much time.  Due to the time commitment and resource
limitation, documentation is rarely emphasized in SCS.

One promising approach to gain the sustainability benefits of documentation, but
significantly reduce the time commitment, is to use a Model Driven Engineering
(MDE) process.  MDE is an engineering approach that exploits structured
representations (metamodels for abstract syntax, and constraints to capture
static semantics) to automate repetitive, tedious and error prone tasks.  In
this research proposal, we will use MDE as a short-hand to include knowledge
models, model transformation, Domain Specific Languages (DSLs) and code
generation.  Several current papers point to DSLs and code/document generation
as a transformative technology for documentation, design and verification of SCS
\citep{JohansonAndHasselbring2018, Smith2018}.  MDE has the potential to
dramatically reduce the cost of developing reliable, reproducible and
re-certifiable software.

One example of applying MDE to SCS is Drasil, which is introduced in
\citet{SzymczakEtAl2016} and available at
\href{https://github.com/JacquesCarette/Drasil}
{https://github.com/JacquesCarette/Drasil}.  Drasil is implemented as an open
source set of Domain Specific Languages (DSLs).  The Drasil approach involves:
i) creating infrastructure for a knowledge base of scientific and computing
models; and, ii) writing explicit ``recipes'' that weave together this knowledge
to generate theoretical models, design documents, code, test cases and build
scripts. The generator can render to multiple languages, such as html, and LaTeX
for documentation and Python, Java, C++ and Lua, for code. One source of
knowledge, with rules for transformation, means completeness, consistency and
traceability can be achieved by construction. Moreover, these qualities can be
maintained as requirements are modified, design decisions are changed,
documentation standards are varied, and software is re­certified.  Although
creating the knowledge base is time consuming, the knowledge can be built up
incrementally.  New projects will reuse existing knowledge and expand as
necessary.

As described above, sustainability is a concern for SCS; MDE holds promise
for addressing this concern.  However, a transformative change in the
development of SCS requires more than promise.  Empirical evidence is necessary
to demonstrate that an MDE process can improve the sustainability of SCS.  This
leads to the long-term objective for this research proposal: \emph{assess and
  measure the impact of MDE on the sustainability of SCS.}

The scope of SCS software for this proposal includes long-lived software, as
implied by the sustainability objective.  To gain the full benefits of
generating documentation, the scope also includes software that interests
multiple stakeholders, with different interests and backgrounds.  Finally, the
scope of this proposal emphasizes safety related software, such as software for
nuclear safety analysis, medical imaging and computational medicine.  Safety
related software is prioritized because documentation is often expected as part
of a certification exercise.  Although this study is not specifically restricted
to open-source software, for practical reasons (especially the lack of access to
the code), commercial software will not be strongly emphasized.

Given the large number of potential MDE processes, techniques and technologies,
together with the wide variety in purpose, scope and context for SCS, the
long-term objective of measuring the impact of MDE is decomposed into four
short-term objectives.  For each objective, a reference is given to the
corresponding section where details are provided.

\begin{enumerate}
\item Assess the current state of the software development practice for SCS.
  How is the software developed in different sub-domains of SCS?  Are there
  existing development methods that lead to higher software sustainability?
  (Section~\ref{SecStateOfThePractice})
\item Assess the impact of an MDE process on end user developed SCS with respect
  to developer productivity and software sustainability for long-lived lived software.  
  (Section~\ref{SecImpactMDE})
\item Assess the impact on safety when the MDE generated documentation targets a
  safety assurance case.  An assurance case is an explicit structured argument
  pertaining to specific properties, such as trustworthiness.  The evidence for
  the argument will come from the traceability between artifacts generated by
  model transformations, expert reviews, test cases, etc.
  (Section~\ref{SecAssuranceCase})
\item Assess the impact on software quality when using MDE to facilitate
  Computational Variability Testing (CVT).  CVT uses code generation to build
  confidence in the generated code in an analogous way to the use of grid
  refinement studies.  Grid refinement looks at how the solution changes by
  varying the run-time parameter of grid density and comparing the results to
  theoretical expectations.  CVT, on the other hand, can generate code to
  ``refine'' build time parameters, such as order of interpolation, or degree of
  implicitness.  These parameters can be systematically varied and the results
  compared against the expected trend.  (Section~\ref{SecCVT})
\end{enumerate}

@@

To assess the impact of MDE on SCS quality, we need a clear definition of what
we mean by quality.  Section~\ref{SecSoftwareQuality} shows how the concept of
quality is decomposed into a set of separate qualities.  This set of qualities
can be applied to the software artifacts (documentation, test cases, etc) and to
the software development process itself.  Several of the qualities from the list
in Section~\ref{SecSoftwareQuality} cannot be measured directly, such as
maintainability.  Therefore, Section~\ref{SecDesirableQs} introduces measurable
documentation qualities that are believed to contribute to the overall software
qualities.  In some cases it may be necessary to use the qualities in
Section~\ref{SecDesirableQs} to indirectly measure qualities listed in
Section~\ref{SecSoftwareQuality}.

The other subsection in this background section provide an overview of the
literature on the current approaches to developing SCS software
\ref{SecCurrStateOfThePractice}, recommended document artifacts for SCS
\ref{SecDocumentation}, an overview of MDE and Code Generation in an SCS context
\ref{SecMDE}, and a summary of existing literature on empirical methods for
software engineering \ref{SecEmpirical}.

@@

\item [\textbf{Sustainability}] A combination of other qualities, likely
  maintainability and productivity.  Reference authorless paper found on-line.

@@

To achieve the qualities listed in Section~\ref{SecSoftwareQuality}, the
documentation should achieve the qualities listed in this section.  All but the
final quality listed (abstraction), are adapted from the IEEE recommended
practise for producing good software requirements~\cite{IEEE1998}.  Abstraction
means only revealing relevant details, which in a requirements document means
stating what is to be achieved, but remaining silent on how it is to be
achieved.  Abstraction is an important software development principle for
dealing with complexity~\cite[p.~40]{GhezziEtAl2003}.
\citet{SmithAndKoothoor2016} present further details on the qualities of
documentation for SCS.

@@

Although reproducibility is the cornerstone of the scientific method, until
recently it has not been treated seriously in software
\citep{BenureauAndRougier2017}.  Fortunately, in recent years multiple
conferences, workshops and individuals are calling for dramatic change
\citep{BaileyEtAl2016}.  The need for action is highlighted by a study of 402
computer systems papers - only 48.3\% of the code was both available and
compilable \citep{CollbergEtAl2015}.  (Drasil addresses this problem because as
programming languages evolve the code renderers in Drasil can be updated.)
Reproducibility problems are even more extreme when the goal is replicability.
A third party should be able to repeat a study using only the description of the
methodology from a published article, with no access to the original code or
computing environment \citep{BenureauAndRougier2017}.  However, replicability is
rarely achieved, as shown for microarray gene expression
\citep{IoannidisEtAl2009} and for economics modelling
\citep{IonescuAndJansson2013}.  Drasil addresses completeness and ambiguity
problems, since it emphasizes capturing and documenting all of the required
knowledge, including derivation of equations and
rationales. \citet{CrickAndHall2014} point out potential roadblocks for
reproducibility, including page length constraints and differing detail needs
depending on the audience.  Again Drasil addresses these concerns because the
recipes used to generate the documentation can be tailored to the level of
detail required.

@@

Thankfully SCS leaders recognize that an interdisciplinary approach provides the
path forward.  They believe that the solution to SCS quality problems is
applying, adapting and developing SE methods, tools and techniques.  However,
typical software processes are a barrier to progress.  ``To break the gridlock,
we must establish a degree of cooperation and collaboration with the [SE]
community that does not yet exist'' \citep{FaulkEtAl2009}.  ``There is a need to
improve the transfer of existing practices and tools from ... [SE] to scientific
programming. In addition, ... there is a need for research to specifically
develop methods and tools that are tailored to the domain'' \citep{Storer2017}.
This tailored research will require teams with a multidisciplinary background.

@@

\wss{Add figure showing the V-model of the documentation.}

@@

DSLs and code/document generation provide a transformative technology for
documentation, design and verification \citep{JohansonAndHasselbring2018,
  Smith2018}.  DSLs allow scientists their preferred approach of focusing on
science not software \citep{Kelly2007}.  A generative approach removes the
maintenance nightmare of documentation duplicates and near duplicates
\citep{LucivEtAl2018}, since knowledge is only captured once and automatically
transformed as needed.  Code generation has previously been applied to improve
SCS \citep{WhaleyEtAl2001, Veldhuizen1998, Pueschel2001}.  For instance, ATLAS
(Automatically Tuned Linear Algebra Software) \citep{WhaleyEtAl2001} and Blitz++
\citep{Veldhuizen1998} produces efficient and portable linear algebra software.
Spiral \citep{Pueschel2001} uses software/hardware generation for digital signal
processing.  \citet{Carette2008} shows how to generate a family of efficient,
type-safe Gaussian elimination algorithms.  FEniCS (Finite Element and
Computational Software) \citep{LoggEtAl2012} uses code generation when solving
differential equations.  Unlike previous work on SCS code generation, Drasil
(\href{https://github.com/JacquesCarette/Drasil}
{https://github.com/JacquesCarette/Drasil}) focus on generating all software
artifacts (requirements, design etc.), not just code.

Drasil, as presented in \citet{SzymczakEtAl2016} removes excuses for avoiding
documentation by providing transformative SCS development technology.  Drasil
provides an infrastructure for knowledge capture and document/code generation.
Drasil is implemented via Domain Specific Languages (DSLs) embedded in Haskell.
Drasil generates requirements documentation and code for several case studies,
including simulating the temperature of a solar water heating system, glass
breakage and two-dimensional game physics.  The full documentation
(requirements, design etc), code and test cases have been created manually for
each case study.  These manual case studies (many from
\citet{SmithJegatheesanAndKelly2016}) provide the `gold standard' against which
Drasil is tested.  All of these artifacts are available publicly.  The
generation techniques for Drasil began to take shape during work on generating
geometric data structures and functions \citep{CaretteEtAl2011}.

@@

The empirical methods in this study will follow the guidelines given in
\citet{KitchenhamEtAl2002}.  Since case studies will be part of the methodology,
the guidelines for conducting and reporting case study research in software
engineering \citep{RunesonAndHost2009} will be followed.

@@

Rule 4: Keep knowledge up to date and findable \citep{ShollerEtAl2019}.

@@

Scientists generally use an approximation of agile methods for their software
development.  In addition, documentation is generally not emphasized.  How
should Drasil be fit into the scientific software development process?  The
Drasil process should encourage an early intensity of thought and effort, but at
the same time, it shouldn't require waterfall development.  Drasil supports
frequent chance and incremental development, but where should developers start
and how should the work proceed?  How does one create an almost empty project in
Drasil and then borrow from existing knowledge and create new knowledge?  This
project could start with something like the
\href{https://jacquescarette.github.io/Drasil/examples/Projectile/srs/Projectile_SRS.html}
{Projectile} example currently in Drasil.

@@

Experimentally Determine Appropriate Documentation Template: The best approach
for documenting requirements and design is an open question.  Drasil could be
used to attempt to address this question.  Different recipes for documentation
could be developed and compared.  Some comparison could be automated between the
documents, such as empirical measures of their quality.  Other measures would
require some kind of usability experiments.
